The Unicode Cookbook \linebreak for Linguists

### 2017-02-02 book/chapters/implementation.tex
updates book for initial public release

\section{Available implementations}
\label{available-implementations}

To illustrate the practical application of orthography profiles as defined in
the previous Chapter \ref{orthography-profiles}, we have implemented two
different versions of the specifications presented there: one in Python and one
in R. These two implementations have rather different implementation histories,
and we are not 100\% sure that they will in all situations give the same
results. Also the perfomance with larger datasets differs, and the code is not
always as clean as we would like it to be. In sum, the two implementations
should be considered as `proof of concept' and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

This chapter will introduce the implementations, and give practical step-by-step 
guidelines for installing them and using them. Various simple and sometimes 
somewhat abstract examples will be discussed to show the different options 
available, and to illustrate the intended usage of orthography profiles. In the 
next chapter we will then discuss a few full-scale examples to illustrate the 
application and the usefulness of orthography profiles.

\subsection*{Installing the R implementation}

@@

TO DO

\subsection*{Installung the Python implementation}

TO DO

\section{Profiles and error reporting}
\label{error-reporting}

@@

\section{Different ways to write a profile}
\label{write-profile}

@@

Note that these different profiles can also be produced using the bash
executable \texttt{writeprofile} (see \ref{available-implementations} for notes
on how to install the bash executable). Exactly this example is also included in
the help file of the executable.

@@

\section{Using an orthography profile skeleton}
\label{profile-skeleton}

@@

\section{Rule ordering}
\label{rule-ordering}

@@

\section{Contextually specified graphemes}
\label{contextual-specification}

@@

\section{Profile skeleton with columns for editing}
\label{profile-editing}

@@

\section{Formatting grapheme separation}
\label{formattingseparation}

@@

\section{Remaining issues}

>>>>

To illustrate the practical application of orthography profiles as defined in
the previous Chapter \ref{orthography-profiles}, we have implemented two 
versions of the specifications presented there: one in Python and one
in R.

These two libraries have rather different implementation histories,
and we are not 100\% sure that they will in all situations give the same
results. We do however provide a test suite for each implementation. 
Users should refer to the tests cases and to the documentation in each release 
for specifics about each implementation. Note also that function names differ 
in the two implementations.

Note also that the performance with larger datasets differs, and the code is not
always as clean as we would like it to be. In sum, the two implementations
should be considered as `proof of concept' and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

This chapter will introduce the implementations, and give practical step-by-step 
guidelines for installing them and using them. Various simple and sometimes 
somewhat abstract examples will be discussed to show the different options 
available, and to illustrate the intended usage of orthography profiles.

This chapter is split into two sections, one for the Python implementation, the 
other for R. Each section has a number of subsections that ...

In addition to the material presented here to get users started, we maintain 
several case studies that use each OP implementation in action. These use cases 
are an the accompanying Github repository available with this book at: 
\url{https://github.com/unicode-cookbook/recipes/}.

% TODO: point to the case studies
% In the next chapter we will then discuss a few full-scale examples to illustrate the application and the usefulness of orthography profiles.

\section{Python}
\label{python-implementations}

The Python package is available both as a command line interface (CLI) 
and as an appplication programming interface (API).

\subsection*{Installation}

To install the Python package from the Python Package Index (PyPI) run:

\begin{lstlisting}[language=bash]
  $ pip install segments
\end{lstlisting}

\noindent on the command line. This will give you access to both 
the CLI and programmatic functionality in Python scripts, when 
you import the \texttt{segments} library.

You can also install the \texttt{segments} 
package with from the Github repository:

\begin{lstlisting}[language=bash]
  $ git clone https://github.com/bambooforest/segments.git
  $ cd segments
  $ python setup.py develop
\end{lstlisting}

\subsection*{Command line interface}

From the command line, access \texttt{segments} and its 
various arguments. For help, run:

\begin{lstlisting}[language=bash]
  $ segments -h
\end{lstlisting}

\noindent Here is an example of how to create and use an orthography 
profile for segmentation. Create a text file:

\begin{lstlisting}[language=bash]
  $ more text.txt
  aäaaöaaüaa
\end{lstlisting}

\noindent Now look at the profile:

\begin{lstlisting}[language=bash,texcl=true]
  $ cat text.txt | segments profile
  Grapheme        frequency       mapping
  a       7       a
  ä       1       ä
  ü       1       ü
  ö       1       ö
\end{lstlisting}

\noindent Write the profile to a file:

\begin{lstlisting}[language=bash]
  $ cat text.txt | segments profile > profile.prf
\end{lstlisting}

\noindent Edit the profile:

\begin{lstlisting}[language=bash]
  $ more profile.prf
  Grapheme        frequency       mapping
  aa      0       x
  a       7       a
  ä       1       ä
  ü       1       ü
  ö       1       ö
\end{lstlisting}

\noindent Now tokenize the text without profile:

\begin{lstlisting}[language=bash]
  $ cat text.txt | segments tokenize
  a ä a a ö a a ü a a	
\end{lstlisting}

\noindent And with profile:
\begin{lstlisting}[language=bash]
  $ cat text.txt | segments --profile=profile.prf tokenize
  a ä aa ö aa ü aa

  $ cat text.txt | segments --mapping=mapping --profile=profile.prf tokenize
  a ä x ö x ü x
\end{lstlisting}

\subsection*{Application programming interface}
The \texttt{segments} API can be accesses by importing 
the package into Python a script. Here is an example of 
how to import the libraries, create a tokenizer object, 
tokenize a string and create a profile.

\begin{lstlisting}[language=python]
>>> from __future__ import unicode_literals, print_function
>>> from segments import Profile, Tokenizer
>>> t = Tokenizer()
>>> t('abcd')
'a b c d'
>>> prf = Profile({'Grapheme': 'ab', 'mapping': 'x'}, \
{'Grapheme': 'cd', 'mapping': 'y'})
>>> print(prf)
mapping Grapheme
ab  x
cd  y
>>> t = Tokenizer(profile=prf)
>>> t('abcd')
'ab cd'
>>> t('abcd', column='mapping')
'x y'
\end{lstlisting}

\section{R}
\label{r-implementation}

% \subsection*{Installing the R implementation}
\subsection*{Installation}

@@

% TO DO

\subsection*{Profiles and error reporting}
\label{error-reporting}

@@

\subsection*{Different ways to write a profile}
\label{write-profile}

@@

Note that these different profiles can also be produced using the bash
executable \texttt{writeprofile} (see \ref{r-implemenation} for notes
on how to install the bash executable). Exactly this example is also included in
the help file of the executable.

@@

\subsection*{Using an orthography profile skeleton}
\label{profile-skeleton}

@@

\subsection*{Rule ordering}
\label{rule-ordering}

@@

\subsection*{Contextually specified graphemes}
\label{contextual-specification}

@@

\subsection*{Profile skeleton with columns for editing}
\label{profile-editing}

@@

\subsection*{Formatting grapheme separation}
\label{formattingseparation}

@@

\subsection*{Remaining issues}

### 2017-02-02 book/chapters/introduction.tex
updates book for initial public release

The resulting cumbersome and often illogical structure of writing systems, and
the enormous variability of existing writing systems for the world's languages
is a fact of life scholars have to accept and should try to adapt to as good as
possible. Our plea in this book is a proposal how to do exactly that: formalize 
knowledge about individual writing systems in a form that is both easy to use 
for linguists in daily practice, and at the same time computer-readable for 
automatic processing.

@@

The necessity to computationally use and compare different orthographies most
forcefully arises in the context of language comparison. Concretely, the proper
processing of orthographies and transcriptions becomes crucial for the
development of quantitative methods for language comparison and historical
reconstruction in order to investigate worldwide linguistic variation and to
model the historical and areal processes that underlie linguistic
diversity.\footnote{
cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
In this work, it is crucial to be able to flexibly process across numerous
resources with different orthographies. In many cases even different resources
on the same language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

@@

Before we deal with these proposals we will extensively discuss the theoretical
background on text encoding, on the Unicode standard, and on the International
Phonetic Alphabet. In the remainder of this chapter, we give an extended
introduction to the notion of encoding (Section~\ref{encoding}) and writing
systems, both from a linguistic perspective
(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
Standard to be a breakthrough (and ongoing) development that fundamentally
changed the way we look at writing systems, and we aim to provide here a
slightly more in-depth survey of the many techniques that are available in the
standard. A good appreciation for the solutions that the Unicode Standard also
allows for a thorough understanding of possible pitfalls that one might
encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of the
current state-of-the-art, we discuss the rather problematic marriage of the
International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter
\ref{ipa-meets-unicode}).

The second part of the book (Sections~\ref{orthography-profiles} and
\ref{use-cases}) describes our proposals for how to deal with the Unicode
Standard in the daily practice of (comparative) linguistics. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
issues with actual examples (Chapter~\ref{use-cases}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries.

The following conventions are followed in this paper. All phonemic and phonetic
representations are given in the International Phonetic Alphabet (IPA), unless
noted otherwise \citep{IPA2005}. Standard conventions are used for
distinguishing between graphemic < >, phonemic / / and phonetic [ ]
representations. For character descriptions, we follow the notational
conventions of the Unicode Standard \citep{Unicode2014}. Character names are
represented in small capital letters (e.g.~\textsc{latin small letter schwa})
and code points are expressed as U\emph{+n} where \emph{n} is a four to six
digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.

>>>>

The resulting cumbersome and often illogical structure of writing systems, and
the enormous variability of existing writing systems for the world's languages
is a fact of life scholars have to accept and should try to adapt to as well as
we can. Our plea in this book is a proposal how to do exactly that: formalize 
knowledge about individual writing systems in a form that is both easy to use 
for linguists in daily practice, and at the same time computer-readable for 
automatic processing.

@@

The necessity to computationally use and compare different orthographies most
forcefully arises in the context of language comparison. Concretely, the proper
processing of orthographies and transcriptions becomes crucial for the
development of quantitative methods for language comparison and historical
reconstruction in order to investigate worldwide linguistic variation and to
model the historical and areal processes that underlie linguistic
diversity.\footnote{
Cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
In this work, it is crucial to be able to flexibly process across numerous
resources with different orthographies. In many cases even different resources
on the same language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

@@

Before we deal with these proposals we will extensively discuss the theoretical
background on text encoding, on the Unicode standard, and on the International
Phonetic Alphabet. In the remainder of this chapter, we give an extended
introduction to the notion of encoding (Section~\ref{encoding}) and writing
systems, both from a linguistic perspective
(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
Standard to be a breakthrough (and ongoing) development that fundamentally
changed the way we look at writing systems, and we aim to provide here a
slightly more in-depth survey of the many techniques that are available in the
standard. A good appreciation for the solutions that the Unicode Standard also
allows for a thorough understanding of possible pitfalls that one might
encounter when using it in general (Chapter~\ref{unicode-pitfalls}). Linguists are more often interested in using the Unicode Standard with the International Phonetic Alphabet (IPA). We first provide a history of the development of the IPA and early attempts to encode it electronically (Chapter~\ref{the-international-phonetic-alphabet}) before we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter\ref{ipa-meets-unicode}).

In the second part of the book (Chapters~\ref{orthography-profiles} and
\ref{implementation}) we describe our proposals of how to deal with the Unicode
Standard in the daily practice of (comparative) linguistics. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
issues with actual examples (Chapter~\ref{implementation}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries.

\subsubsection*{Conventions}

The following conventions are followed in this book. All phonemic and phonetic
representations are given in the International Phonetic Alphabet (IPA), unless
noted otherwise \citep{IPA2005}. Standard conventions are used for
distinguishing between graphemic < >, phonemic / / and phonetic [ ]
representations. For character descriptions, we follow the notational
conventions of the Unicode Standard \citep{Unicode2014}. Character names are
represented in small capital letters (e.g.~\textsc{latin small letter schwa})
and code points are expressed as U\emph{+n} where \emph{n} is a four to six
digit hexadecimal number, e.g.~\uni{0256}, which can be rendered as the glyph <ə>.

### 2017-02-02 book/chapters/ipa_background.tex
updates book for initial public release

@@

Established in 1886, the \textsc{international phonetic association} (henceforth
\textit{Association}) has long maintained a standard alphabet, the
\textsc{international phonetic alphabet} (henceforth \textit{IPA}), which is a
common standard in linguistics to transcribe sounds of spoken languages. It was
first published in 1888 as an international system of phonetic transcription for
oral languages and for pedagogical purposes. It contained phonetic values for
English, French and German. Diacritics for length and nasalization were already
present in this first version, and the same symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

Originally, the IPA was a list of symbols with pronunciation examples
using words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels (e.g.\ for manner of
articulation among others `plosives', `nasales', `fricatives', for place of
articulation among others `bronchiales', `laryngales', `labiales' and for vowels
e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, this is part of the
Association's mandate. These changes are captured in the recurring revisions to
the IPA.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there have been several revisions, many minor. For example, the
articulation labels -- what often are call `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The current construction and use of the IPA are guided by
principles outlined in the \textsc{handbook of the international phonetic
association} \citep{IPA1999}, henceforth simply called \textit{Handbook}.

@@

Several styles of transcription with the IPA are possible, as illustrated in the
\textit{Handbook}, and they are all valid.\footnote{For illustration of
different possibilities, see the 29 languages and their transcriptions in the
original and initial \textit{Illustrations of the IPA}
\citep[41--154]{IPA2007}} Therefore, there are different but equivalent
transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``perhaps now that
the Association has been explicit in its eclectic approach, outsiders to the
Association will no longer speak of \textit{the} IPA transcription of a given
phenomenon, as if there were only one approved style.'' Clearly not all
phoneticians agree (or will likely ever agree of course) on all aspects of the
IPA or on transcription practices. As noted above, there have been several
revisions in the IPA's long history, but the current version (2005) is
strikingly similar to the 1926 version, showing the viability of the IPA as a
common standard for linguistic transcription.

% ==========================
\section{Premises}
\label{IPApremises}
% ==========================

@@

% ==========================
\section{Principles}
\label{IPAprinciples}
% ==========================

Transcription has essentially two parts. The first is a text containing IPA 
symbols and the second is a set of conventions for interpreting those 
symbols (and their combinations).

First, the symbols of IPA crucially distinguish between letter-like symbols and
diacritics. The use of the letter-like symbols to represent a language's sounds
is guided by the principle of contrast; where two words are distinguishable by
phonemic contrast, those contrasts should be transcribed with different letter
symbols (not diacritics). Allophonic distinction falls under the rubric of
diacritically-distinguished symbols, e.g. the difference in English between and
unrealized /p/ in [stop̚] vs. an aspirated /p/ in [spʰot] should be marked with
a diacritic. In sum:

@@

	\item Different letter-like symbols should be used whenever
          a language employs two sound contrastively.
	\item When two sounds in a language are not known to be contrastive, the same
          symbol should be used to represent these sounds. Diacritics may
          be used to distinguish different articulations when necessary.
\end{itemize}

\noindent Yet, in some situations diacritics can be used to mark phonemic
contrasts. The \textit{Handbook} recommends to limit the use of phonemic
diacritics to the following situations:

@@

% ==========================
\section{Computational representation}
% ==========================

In 1989 an IPA revision convention was held in Kiel, Germany. As per other
revisions, there was expansion and revisions to phonetic symbols in the IPA
chart. Notably the marking of tone was extended with the addition of a second
system for marking linguistic tones (the so-called `Chao' tone letters).
Importantly, however, this was also the first revision to address issues of
computational representations for the IPA.

Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized ``computer
codes'' to phonetic symbols was assigned to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

The working group insightfully decided that the computing-coding convention 
for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
i.e.\ the IPA Number was not meant to be implemented directly in a computer 
encoding. The working group report's declaration includes the following explanatory 
remarks \citep[82]{International1989report}:

@@

\noindent Thus the coding convention was not meant to address the engineering 
aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
it was meant to serve as a basis for a communication-interchange standard for 
creating mapping tables from various computer encodings, fonts, phonetic-character-set 
software, etc., to common IPA Numbers, and thus symbols.\footnote{Remember, at 
this time in the late 1980s there was no stable multilingual computing environment. 
But some solution was needed because scholars were increasingly using personal 
computers for their research and many were quickly adopting electronic mail or 
discussion boards like Usenet as a medium for international exchanges. Most of these systems ran on 8-bit 
hardware systems using a 7-bit ASCII character encoding.}

@@

After a symbol is categorized, it is assigned a number sequentially, e.g.\ [i]
301, [e] 302, [ɛ] 303. The system allows for the addition of new symbols within
the various series by appending them, e.g. [\charis{ⱱ}] 184. Former non-IPA
symbols (or often-used but non-offical IPA symbols) for consonants, vowels and
diacritics are numbered from x99 backwards. For example, the voiceless and
voiced postalveolar affricates and fricatives <~č,~ǰ,~š,~ž~> are assigned the
IPA numbers 299, 298, 297 and 296, respectively, because they are not sanctioned
IPA symbols.

@@

% ==========================
\section{SAMPA and X-SAMPA}
% ==========================

@@

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A more recent survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990}. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

@@

% ==========================
\section{The need for a single multilingual environment}
% ==========================

@@

% ===============================================================

\begin(comment)

\subsection{IPA (practical) extensions}

The IPA extensions cover speech sounds beyond the sound systems of the world's
languages, including symbols for pathological speech and paralinguistic
functions (e.g.\ Braille?).

- pathological speech
- paralinguistic functions (Braille?)

As noted in Section \ref{}, extensions to the IPA -- in particular
``large-scale'' extensions --

% Englebretson2009.pdf
http://www.ruf.rice.edu/~reng/englebretson2009.pdf

X-SAMPA was an ASCII transliteration and is still purposeful for several
programs

There is also the extensions of the IPA designed for disordered speech, although
the symbols are occasionally used for transcribing phonetic detail in ``normal''
speech or also cross-linguistic
rarities.\footnote{\url{http://phoible.github.io/conventions/}}

% Unicode has its own history; for a summary:
\url{http://www.unicode.org/history/summary.html}

https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm

Computer codes for phonetic symbols
EslingGaylord1993

% ==========================
\section{Unicode and ISO 10646}
% ==========================

In the late 1980s, a universal character set was being developed by what 
is now referred to as the Unicode Consortium (officially incorporated in January 1991).
This consortium 
consisted largely, although not entirely, of major US corporations, with 
the aim of overcoming the inoperability of different coded character sets 
and their costly hinderance for developing multilingual software development 
and for internationalization efforts. Commercial importance of course drove 
the early inclusion of Latin, non-Latin, and some exotic scripts; see the 
table of commercial importance as measured by GDP of countries using certain 
writing systems \citep[2]{unicode88}.

The original Unicode manifesto is \cite{unicode88.pdf}.\footnote{http://www.unicode.org/history/unicode88.pdf} 
Its aim was for a reliable international multilingual text encoding standard 
that would encompass all scripts of the world, or in the author's own words, 
``a new, world-wide ASCII''. An in-depth history of Unicode, highlighting 
interesting facts like its first text prototypes at Apple and its incorporation 
into TrueType, is retold online.\footnote{\url{http://www.unicode.org/history/earlyyears.html}}

>>>>

In this chapter we present a brief history of the IPA 
(Section~\ref{IPAhistory}), which itself dates 
back to the late 19th century, not long after the creation of the first 
typewriter with a QWERTY 
keyboard. An understanding of the IPA and its premises and principles 
(Section~\ref{IPApremises-principles}) leads to a better 
appreciation of the challenges that the IPA association 
faced when digitally encoding the IPA's set of symbols and 
diacritics (Section~\ref{EncodingIPA}). Occurring a bit over one hundred 
years after the IPA's inception, the encoding challenge was no small feat 
(Section~\ref{need-for-multilingual-environment}); many 
linguists have encountered the pitfalls when the twain shall meet 
(Chapter~\ref{ipa-meets-unicode}).

@@

Established in 1886, the \textsc{international phonetic association} (henceforth
\textit{Association}) has long maintained a standard alphabet, the
\textsc{international phonetic alphabet} or IPA, which is a
standard in linguistics to transcribe sounds of spoken languages. It was
first published in 1888 as an international system of phonetic transcription for
oral languages and for pedagogical purposes. It contained phonetic values for
English, French and German. Diacritics for length and nasalization were already
present in this first version, and the same symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

Originally, the IPA was a list of symbols with pronunciation examples
using words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels (e.g.\ for manner of
articulation among others `plosives', `nasales', `fricatives', for place of
articulation among others `bronchiales', `laryngales', `labiales' and for vowels
e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge. Thus, similar to notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, updates are part of the
Association's mandate. These changes are captured in revisions to the IPA chart.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there have been several revisions, but mostly minor. Articulation 
labels -- what are often called `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The use of the IPA is guided by principles outlined in 
the \textsc{handbook of the international phonetic association} \citep{IPA1999}, 
henceforth simply called \textit{Handbook}.

@@

Several styles of transcription with the IPA are possible, as illustrated in the
\textit{Handbook}, and they are all valid.\footnote{For an illustration of
the differences, see the 29 languages and their transcriptions in the
\textit{Illustrations of the IPA}\citep[41--154]{IPA2007}.} Therefore, there are 
different but equivalent transcriptions, or as noted by \citet[64]{Ladefoged1990a}, 
``perhaps now that the Association has been explicit in its eclectic approach, outsiders to the
Association will no longer speak of \textit{the} IPA transcription of a given
phenomenon, as if there were only one approved style.'' Clearly not all
phoneticians agree, nor are they likely to ever completely agree, on all aspects of the
IPA or on transcription approaches and practices in general. As noted above, 
there have been several revisions in the IPA's long history, but the current version (2005) is
strikingly similar to the 1926 version, which shows the viability of the IPA as a
common standard for linguistic transcription.

% ==========================

\section{Premises and principles}
\label{IPApremises-principles}
% ==========================
\subsection*{Premises}
\label{IPApremises}

@@

% ==========================
\subsection*{Principles}
\label{IPAprinciples}
% ==========================

Essentially transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ in [spʰot] and 
an unrealeased /p/ in [stop̚].

@@

	\item Different letter-like symbols should be used whenever
          a language employs two sounds contrastively.
	\item When two sounds in a language are not known to be contrastive, the same
          symbol should be used to represent these sounds. Diacritics may
          be used to distinguish different articulations when necessary.
\end{itemize}

\noindent Yet, in some situations diacritics are used to mark phonemic
contrasts. The \textit{Handbook} recommends to limit the use of phonemic
diacritics to the following situations:

@@

% ==========================

\section{IPA encodings}
\label{EncodingIPA}
% ==========================

In 1989, an IPA revision convention was held in Kiel, Germany. As in previous meetings, 
there were changes made to the repertoire of phonetic symbols in the IPA chart, which 
reflected what had been discovered, described and cataloged by linguists about the 
phonological systems in the world's languages in the interim. Personal computers 
were also becoming more commonplace, and linguists were using them to create databases. 
A cogent example is the UCLA Phonological 
Segment Inventory Database (UPSID; \citealt{Maddieson1984}), which was expanded 
\citep{MaddiesonPrecoda1990} and then encoded and distributed in a computer program 
\citep{MaddiesonPrecoda1992}.\footnote{It could be installed via 
floppy disk on an IBM PC, or compatible, running 
DOS with 1MB free disk space and 360K available RAM.} The programmers used 
only ASCII characters to maximize compatibility (e.g.\ <kpW> for $[$kpʷ$]$), but 
were faced with unavoidable arbitrary mappings between ASCII letters and 
punctuation, and the more than 900 segment types documented 
in their sample of world's languages' phonological systems. The developers 
devised a system of base characters with secondary diacritic marks 
(e.g.\ in the previous example <kp>, the base character, is modified with <W>). 
This encoding approach is 
also used in SAMPA and X-SAMPA (Section~\ref{sampa-xsampa}) and in the 
ASJP.\footnote{See the ASJP use case in the online supplementary 
materials to this book: \url{https://github.com/unicode-cookbook/use_cases}.} 
But before UPSID, SAMPA and ASJP, IPA was encoded with numbers.

% \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).} 
\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized ``computer
codes'' to phonetic symbols was assigned to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

The working group decided insightfully that the computing-coding convention 
for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
i.e.\ the IPA Number was not meant to be implemented directly in a computer 
encoding. The working group report's declaration includes the following explanatory 
remarks \citep[82]{International1989report}:

@@

\noindent The coding convention was not meant to address the engineering 
aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
it was meant to serve as a basis for a communication-interchange standard for 
creating mapping tables from various computer encodings, fonts, phonetic-character-set 
software, etc., to common IPA Numbers, and therefore symbols.\footnote{Remember, at 
this time in the late 1980s there was no stable multilingual computing environment. 
But some solution was needed because scholars were increasingly using personal 
computers for their research and many were quickly adopting electronic mail or 
discussion boards like Usenet as a medium for international exchanges. 
Most of these systems ran on 8-bit hardware systems using a 7-bit ASCII character encoding.}

@@

After a symbol is categorized, it is assigned a number sequentially, e.g.\ [i]
301, [e] 302, [ɛ] 303. The system allows for the addition of new symbols within
the various series by appending them, e.g. [\charis{ⱱ}] 184. Former non-IPA
symbols (or often-used but non-official IPA symbols) for consonants, vowels and
diacritics are numbered from x99 backwards. For example, the voiceless and
voiced postalveolar affricates and fricatives <~č,~ǰ,~š,~ž~> are assigned the
IPA numbers 299, 298, 297 and 296, respectively, because they are not sanctioned
IPA symbols.

@@

% ==========================
\subsection*{SAMPA and X-SAMPA}
\label{sampa-xsampa}
% ==========================

@@

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A more recent survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}.} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990}. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

@@

% ==========================

\section{The need for a single multilingual environment}
\label{need-for-multilingual-environment}
% ==========================

@@

% ==========================

\begin{comment}
\section{Unicode and ISO 10646}
% ==========================

In the late 1980s, a universal character set was being developed by what 
is now referred to as the Unicode Consortium (officially incorporated in January 1991).
This consortium 
consisted largely, although not entirely, of major US corporations, with 
the aim of overcoming the inoperability of different coded character sets 
and their costly hinderance for developing multilingual software development 
and for internationalization efforts. Commercial importance of course drove 
the early inclusion of Latin, non-Latin, and some exotic scripts; see the 
table of commercial importance as measured by GDP of countries using certain 
writing systems \citep[2]{Becker1988}.

The original Unicode manifesto is \cite{Becker1988}.\footnote{http://www.unicode.org/history/unicode88.pdf} 
Its aim was for a reliable international multilingual text encoding standard 
that would encompass all scripts of the world, or in the author's own words, 
``a new, world-wide ASCII''. An in-depth history of Unicode, highlighting 
interesting facts like its first text prototypes at Apple and its incorporation 
into TrueType, is retold online.\footnote{\url{http://www.unicode.org/history/earlyyears.html}}
